---
title: 'P&S-2022: Lab assignment 3'
author: "Sofiia Folvarochna, Anastasiia Shvets, Serhii Ivanov"
output:
  html_document:
    df_print: paged
---

```{r}
id <- 6
set.seed(id)
```

# Problem 3

```{r}
theta <- id/10
alphas <- c(0.01, 0.05, 0.1)
bethas <- 1-alphas/2 
n <- 100
m <- 10000
exp_sample <- matrix(rexp(n*m, 1/theta), nrow = n)
sample_mean <- colMeans(exp_sample)
sample_sd <- apply(exp_sample, 2, sd)
```

The expected value of the exponential distribution $\mathcal{E}(\lambda)$ is$1/\lambda$, so that a good point estimate of the parameter $\theta:= 1/λ$ is the sample mean $\mathbf{\overline{x}}$. Confidence interval for $\theta$ can be formed in several different ways:

(1) Using the exact distribution of the statistics $2λn\mathbf{\overline{X}}$ (show it is ${\chi}^2_{2n}$ and then use quantiles of the latter to get the interval endpoints)

    $$
    2\lambda n \mathbf{\overline{X}} = \frac{X_1+...+X_n}{n}2\lambda n = 2\lambda X_1+...+2\lambda X_n 
    $$

    $$
    F_{2\lambda X}(t) =P(2\lambda X<t) = P(X<\frac{t}{2\lambda}
    ) = F_X(\frac{t}{2\lambda})$$

    $$
    (F_X(\frac{t}{2\lambda}))' = f_x(\frac{t}{2\lambda})\frac{1}{2\lambda}) = \lambda e^{-\lambda\frac{t}{2\lambda}}\frac{1}{2\lambda}=\frac{1}{2}e^{-\frac{t}{2}}\sim\mathcal{E}(\frac{1}{2})
    $$

    ::: {align="center"}
    $2\lambda n\mathbf{\overline{X}} = Y_1 + ...Y_n$, where every $Y_i \sim \mathcal{E}(\frac{1}{2})$, so $2\lambda n\mathbf{\overline{X}}\sim \Gamma(n, \frac{1}{2})$.
    :::

    $$
    P(\chi^{2n}_\frac{\alpha}{2} \le
     2\lambda n\mathbf{\overline{X}} \le  \chi^{2n}_{1-\frac{\alpha}{2}})$$

    $$
    P(\frac{\chi^{2n}_\frac{\alpha}{2}}{2n\mathbf{\overline{X}}} \le \lambda \le \frac{\chi^{2n}_{1-\frac{\alpha}{2}}}{2n\mathbf{\overline{X}}}
    $$

    $$P(\frac{2n\mathbf{\overline{X}}}{\chi^{2n}_{1-\frac{\alpha}{2}}} \le \theta \le \frac{2n\mathbf{\overline{X}}}{\chi^{2n}_\frac{\alpha}{2}})$$

    $$
    \left[
    \begin{array}{ccc}
    \frac{2n\mathbf{\overline{X}}}{\chi^{2n}_{1-\frac{\alpha}{2}}} & \frac{2n\mathbf{\overline{X}}}{\chi^{2n}_\frac{\alpha}{2}}
    \end{array}
    \right]
    $$

```{r}
for (alpha in alphas){
    cat("For confidence level", 1- alpha, "\n")
    cat("    the fraction of CI's containing the parameter is",
        mean((theta >= (2*n*sample_mean)/qchisq(1-alpha/2, 2*n)) & (theta <= (2*n*sample_mean)/qchisq(alpha/2, 2*n))), "\n", sep = " ")
    cat("    maximal CI length is", (2*n*max(sample_mean))/qchisq(alpha/2, 2*n) - (2*n*max(sample_mean))/qchisq(1-alpha/2, 2*n), "\n", sep = " ")
    cat("    mean CI length is", (2*n*mean(sample_mean))/qchisq(alpha/2, 2*n) - (2*n*mean(sample_mean))/qchisq(1-alpha/2, 2*n), "\n", sep = " ")
}
```

(2) Using the normal approximation $\mathcal{N}(\mu,\sigma^2)$ for $\mathbf{\overline{X}}$; the parameters are $\mu$ = $\theta$ and $\sigma^2$ = $s^2/n,$ where $s^2$ = $\theta^2$ is the population variance (i.e., variance of the original distribution $\mathcal{E}(λ)$). In other words, we form the Z-statistics $Z$ := $\sqrt{n}(\mathbf{\overline{X}} - \theta)/\theta$ and use the fact that it is approximately standard normal $\mathcal{N}(0, 1)$ to find that

    $$P(|\theta - \mathbf{\overline{X}}| \le z_\beta\theta/\sqrt(n) = P(|Z| \le z_\beta) = 2\beta - 1$$

    in other words, $\theta$ is with probability $2\beta - 1$ within $\mathbf{\overline{X}}\pm z_\beta\theta/\sqrt(n)$ .

```{r}



for (alpha in alphas){
    cat("For confidence level", 1- alpha, "\n")
    cat("    the fraction of CI's containing the parameter is",
        mean((theta >= (sample_mean - qnorm(1 - alpha/2)*theta/sqrt(n))) & (theta <= (sample_mean + qnorm(1 - alpha/2)*theta/sqrt(n)))), "\n", sep = " ")
    cat("    maximal CI length is", (max(sample_mean) + qnorm(1 - alpha/2)*theta/sqrt(n)) - (max(sample_mean) - qnorm(1 - alpha/2)*theta/sqrt(n)), "\n", sep = " ")
    cat("    mean CI length is", (mean(sample_mean) + qnorm( 1 - alpha/2)*theta/sqrt(n)) - (mean(sample_mean) - qnorm(1 - alpha/2)*theta/sqrt(n)), "\n", sep = " ")
}
```

3.  The confidence interval constructed above uses the unknown variance $s^2=\theta^2$ and is of little use in practice. Instead, we can solve the double inequality

    $$|\theta - \mathbf{\overline{X}}| \le z_\beta\theta/\sqrt{n}$$

    $$
    \begin{cases}
    \theta \le \frac{\mathbf{\overline{X}}}{1-\frac{z_\beta}{\sqrt{n}}}\\
    -\theta \le \frac{-\mathbf{\overline{X}}}{1+\frac{z_\beta}{\sqrt{n}}}  
    \end{cases}
    $$

    $$
    \begin{cases}
    \theta \le \frac{\mathbf{\overline{X}}}{1-\frac{z_\beta}{\sqrt{n}}}\\
    \theta \ge \frac{\mathbf{\overline{X}}}{1+\frac{z_\beta}{\sqrt{n}}}  
    \end{cases}
    $$

    $$
    \left[
    \begin{array}{ccc}
    \frac{\mathbf{\overline{X}}}{1+\frac{z_\beta}{\sqrt{n}}} & \frac{\mathbf{\overline{X}}}{1-\frac{z_\beta}{\sqrt{n}}}
    \end{array}
    \right]
    $$

```{r}
for (alpha in alphas){
    cat("For confidence level", 1- alpha, "\n")
    cat("    the fraction of CI's containing the parameter is",
        mean((theta >= (sample_mean/(1+ qnorm(1 - alpha/2)/sqrt(n)))) & (theta <= (sample_mean/(1- qnorm(1 - alpha/2)/sqrt(n))))), "\n", sep = " ")
    cat("    maximal CI length is",(max(sample_mean)/(1- qnorm(1 - alpha/2)/sqrt(n))) - (max(sample_mean)/(1+ qnorm(1 - alpha/2)/sqrt(n))), "\n", sep = " ")
    cat("    mean CI length is", (mean(sample_mean)/(1- qnorm(1 - alpha/2)/sqrt(n))) - (mean(sample_mean)/(1+ qnorm(1 - alpha/2)/sqrt(n))), "\n", sep = " ")
}
```

4.  Another (and a more universal approach) to get rid of the dependence on $θ$ in (2) is to estimate s via the sample standard error and use approximation of $\mathbf{\overline{X}}$ via Student t-distribution; see details in Ross textbook on statistics or in the lecture notes

    $$
    \frac{\mathbf{\overline{X}}-\theta}{\sigma}\sqrt{n} \sim \mathcal{N}(0,1)
    $$

    $$
    \frac{\mathbf{\overline{X}}-\theta}{S}\sqrt{n} \sim \mathcal{T}_{n-1}\\
    P(t^{n-1}_{\frac{\alpha}{2}} \le \frac{\mathbf{\overline{X}}-\theta}{S}\sqrt{n} \le t^{n-1}_{1-\frac{\alpha}{2}}) \\
    P(\mathbf{\overline{X}} -  t^{n-1}_{1-\frac{\alpha}{2}} \frac{S}{\sqrt{n}} \le \theta \le \mathbf{\overline{X}}  + t^{n-1}_{1-\frac{\alpha}{2}} \frac{S}{\sqrt{n}})\\
    \left[
    \begin{array}{ccc}
    \mathbf{\overline{X}} -  t^{n-1}_{1-\frac{\alpha}{2}} \frac{S}{\sqrt{n}} & \mathbf{\overline{X}}  + t^{n-1}_{1-\frac{\alpha}{2}} \frac{S}{\sqrt{n}}
    \end{array}
    \right]
    $$

```{r}
for (alpha in alphas){
    cat("For confidence level", 1- alpha, "\n")
    cat("    the fraction of CI's containing the parameter is",
        mean(abs(sample_mean -theta) < qt(1-alpha/2, n - 1)*sample_sd/sqrt(n)), "\n", sep = " ")
    cat("    maximal CI length is", 2*(qt(1-alpha/2, n - 1)*max(sample_sd)/sqrt(n)), "\n", sep = " ")
    cat("    mean CI length is", 2*(qt(1-alpha/2, n - 1))*mean(sample_sd)/sqrt(n),"\n", sep = " ")
}
```
## Conclusions

In my opinion, chi-squared first method is the best for forming confidence intervals for exponential distribution., as it gives the closest fraction of CI's containing the parameter.

Second and third method are dependent on $\theta$, as they use the same inequality with unknown variance, so I wouldn't recommend to use them.

Fourth method gives less accurate results, so that's why first method is the best for this task.

The biggest mean CI length gives the third method, which is unexpected for me, but with bigger n and m the difference between CI lengths calculated using different methods becomes really small.

I also found interesting that results are more accurate with bigger M(amount of samples), and they are less dependent on N(length of sequence).
